\documentclass[a4paper,USenglish,cleveref,autoref]{lipics-v2021}

%% Regular papers for ITP 2024 must:
% be no more than 16 pages in length excluding bibliographic references
% not include an appendix; and
% be in LIPIcs format.
% Submissions will undergo single-blind peer review. 

%This is a template for producing LIPIcs articles. 
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
\usepackage{microtype}%if unwanted, comment out or use option "draft"
%\pdfoutput=1
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathpartir}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{minted}
\usepackage{url}
\usepackage{graphicx}
\usepackage{stmaryrd}
\usepackage{latex/agda}
\usepackage{comment}
\usepackage{todonotes}
\newcommand\pt{\todo[author=PT,inline]}
\newcommand\atha{\todo[author={@Ha},inline]}
\lstdefinelanguage{michelson}{
  basicstyle=\fontsize{8}{9.6}\selectfont,
  morekeywords={parameter,storage,or,unit,mutez,pair,bool,address}, sensitive=false,
  morecomment=[l]{\#},
  morecomment=[\STACK]{/*}{*/},
  morestring=[b]",
}
\lstset{
  language=Caml,
  captionpos=b,
  aboveskip=-\smallskipamount,
  belowskip=-\smallskipamount,
  belowcaptionskip=0pt,
  basicstyle=\fontsize{8}{9.6}\selectfont,
  morekeywords={val}
}

\input{latex/agda-generated}
\input{unicodeletters}
\input{macros}


\bibliographystyle{plainurl}% the mandatory bibstyle

\title{A Dynamic Logic for the Smart Contract Programming Language Michelson}
% \title{Verifying Symbolic Execution for the Smart Contract Programming
% Language Michelson} %TODO Please add

%\titlerunning{Dummy short title} %TODO optional, please use if title is longer than one line

\author{Barnabas Arvay}{University of Freiburg, Germany}{barnabasarvay@gmail.com}{https://orcid.org/0009-0002-2720-7100}{}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

\author{Peter Thiemann}{University of Freiburg, Germany}{thiemann@informatik.uni-freiburg.de}{https://orcid.org/0000−0002−9000−1239}{}

\authorrunning{B. Arvay and P. Thiemann} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Barnabas Arvay and Peter Thiemann} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10010940.10010992.10010998.10011000</concept_id>
<concept_desc>Software and its engineering~Automated static analysis</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Automated static analysis}

\keywords{Smart Contract, Blockchain, Formal Verification, Symbolic Execution} %TODO mandatory; please add comma-separated list of keywords

%\category{} %optional, e.g. invited paper

%\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

%\acknowledgements{I want to thank \dots}%optional

%\nolinenumbers %uncomment to disable line numbering



%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\maketitle

%TODO mandatory: add short abstract of the document
\begin{abstract}
Verification of smart contracts is an important topic in the context
of blockchain technology. We study an approach to verification that is
based on symbolic execution. 

As a formal basis for symbolic execution, we design a dynamic logic
for Michelson, the smart contract language of the Tezos blockchain,
and prove its soundness in the proof assistant Agda. Towards the
soundness proof we formalize the concrete semantics as well as its
symbolic counterpart in a unified setting. The logic encompasses
single contract runs as well as several runs chained in a single transaction. 
\end{abstract}

\input{1-introduction}
\input{2-design-typing}
\input{3-reference-implementation}
\input{4-dynamic-logic}
\input{5-semantics-and-soundness}
\input{6-conclusion}


\section{Related Work}
\label{sec:related-work}

Research on the formal verification of blockchain-based applications has experienced rapid growth in the last decade. Various techniques and frameworks have been applied to enhance the security and reliability of smart contracts on the blockchain. In this section, we discuss some key approaches, particularly those employing symbolic execution in the context of smart contracts.
 
Symbolic execution has emerged as a powerful technique for
systematically exploring program paths and identifying potential
vulnerabilities in smart contracts. P. Tsankov et al. introduced
SECURIFY \cite{securify}, a tool that utilizes symbolic execution to
perform practical security analysis on Ethereum smart contracts. It
targets common vulnerability security patterns specified in a
designated domain-specific language. SECURIFY symbolically encodes the
dependence graph of the contract in stratified Datalog to extract
semantic information from the code. After obtaining semantic facts, it
checks whether the security patterns hold or not. Similarly, Manticore
\cite{manticore} and KEVM \cite{kevm} use symbolic execution to
analyze Ethereum smart contracts. KEVM is an executable formal
specification built with the K Framework for the Ethereum virtual
machine's bytecode (EVM), a stack-based and low-level smart contract
language for the Ethereum blockchain. Since tokens can hold a
significant amount of value, they are often targeted for
attacks. Therefore, several tools \cite{kevm,park} conduct case
studies for the implementations of token standards. 

Several approaches use existing formal verification frameworks to
ensure the correctness and security of smart contracts. S. Amani et
al. \cite{isabelle} proposed the formal verification of Ethereum smart
contracts in Isabelle/HOL. Hirai \cite{hirai} formalizes the EVM using
Lem, a language to specify semantic definitions. The formal
verification of smart contracts is achieved using the Isabelle proof
assistant. Bernardo et al.~\cite{micho} implemented Mi-Cho-Coq, a
formalization of the semantics of Michelson using the Coq proof
assistant. They also verified several Michelson contracts. 

There are exsiting tools for automated verification include solc-verify \cite{solc}, VerX \cite{verx}, and Oyente \cite{oyente}. solc-verify processes smart contracts written in Solidity and discharges verification conditions using modular program analysis and SMT solvers. It operates at the level of the contract source code, with properties specified as contract invariants and function pre- and post-conditions provided as annotations in the code by the developer. This approach offers a scalable, automated, and user-friendly formal verification solution for Solidity smart contracts. The core of solc-verify involves translating Solidity contracts to Boogie IVL (Intermediate Verification Language), a simple language designed for verification. 

While these approaches differ, they share a focus on common bugs (such as reentrancy, overflow or underflow, and frozen funds). Our tool provide an environment where users can reliably specify their own properties and identify contract-specific bugs. Our tool aims to rich specifications to supports manually specified full correctness specifications.


Close to our approach are HELMHOLTZ \cite{helmholtz} and iContract \cite{icontract}. Y. Nishida et al. \cite{helmholtz} developed HELMHOLTZ, an automated verification tool for Michelson. While both research efforts aim to build a verification tool for smart contracts written in Michelson, HELMHOLTZ is based on the refinement type system, whereas ours relies on symbolic execution. Both tools use the SMT solver Z3 for property verification. HELMHOLTZ takes, as input, a Michelson program annotated with a user-defined specification expressed in a refinement type; it then typechecks the program against the specification using the refinement type system. Similar to our tool, it discharges verification conditions with the SMT solver Z3. If the code successfully typechecks, then the program is guaranteed to satisfy the specification. Both tools use a language for user specifications, with properties annotated inside the code by the developer. Here, we provide a separate specification language that users don't need to modify within the code. While their specification language is close to ML-like notations, our language supports logic formulas in first-order logic similar to Z3. While HELMHOLTZ conducts example case studies or examines some entry points of real contracts, our tool conducts two real case studies of running blockchain contracts, implementing significant subjects on the blockchain such as token standards and oracles.

iContract targets the Solidity language. It also utilizes pre and post conditions, similar to our tool, to specify user properties. It locally installs the Solidity compiler to compile a user-provided Solidity file into a JSON file containing the typed abstract syntax tree (AST). Then, iContract analyzes the AST to encode contracts into predicates using the Z3 library. They leverage the NatSpec format to define their own specifications.

\section{Conclusion}
\label{sec:concl-sect-append}

Symbolic execution is well known to suffer from challenges related to
state explosion, particularly when applied to large
systems. Fortunately, the constraints of storing and running smart
contracts on a blockchain platform impose size limitations, making
them manageable for symbolic execution. 

In future work, we want to extend SCV in two directions. 
First, we want to address the interaction of several entrypoints by
extending the specification language to state allowed sequences of
contract invocations. 

Second, a Michelson program returns a list of instructions, which may contain
further contract invocations. Dealing with this kind of contracts
requires an extension to the symbolic interpreter so that a sequence
of contract invocations can be processed in one run.



\bibliography{ecoop2024,bib/refs}
\appendix
\include{appendix}
\end{document}
